---
title: "Notebook"
output: html_document
date: "2024-11-25"
---

# 1.Set Up

```{r}
rm(list = ls())

# For reproducibility pourposes
set.seed(123)

# List of required libraries
required_libraries <- c("skimr", "ggplot2","dplyr")

# Install and load libraries
for (lib in required_libraries) {
  if (!require(lib, character.only = TRUE)) {
    install.packages(lib, dependencies = TRUE)
    library(lib, character.only = TRUE)
  }
}
```

We will start obtaining the datasets and interpreting what we have as well as making some decision which could be useful for our future analysis. We can see that both training and test datasets are already separated, they have different dimensionality as the training set has the target variable, therefore we cannot merge them. As the total dataset would have around 3000 observations a 50% separation of the total dataset for training and test seems fairly good.

```{r}
training <- read.csv("train.csv", sep = ",")
testing <- read.csv("test.csv", sep = ",")
```

```{r}
skim(training)
```

With a first visualization of the dataset we can see that 43 of the columns are categorical while 38 of them are numerical.

# 2.Data Preprocessing

We can see that there are some columns that have many NA values, we will use a barplot to visualize this better and decide whether we should remove columns or not.

## 2.1 NA Values

```{r}
# Plotting a barplot with the number of NA values
barplot(colMeans(is.na(training)), las=2)
```

Based on the previous plot and the skim function used we can see that there are some variables which have NA values for most if not almost all of the observations.

```{r}
# Finidng out the columns with a high number of NA values
colnames(training)[which(colMeans(is.na(training))>0.75)]
```

These columns are Alley, FirePlaceQu, PoolQC, Fence and MiscFeature

Alley refers to the type of alley access to the property. It can be Gravel or Paved, NA means no alley, almost all houses have an NA value here, we don't think that the type of alley is of great importance we will remove it

PoolQC refers to the quality of the swimming pool, as most houses don't have a pool and there are other variables such as PoolArea which refers to the dimension of the swimming pool, we will remove it.

Fence refers to the quality of the fencing, following a similar explanation as before, it has too many NA values, it won't provide much information

MiscFeature mentions other additional features houses may have or not, such as tennis courts. As this are really exclusive and as with the other variables there are a lot of missing values we will delete it.

```{r}
# Deleting the variables
na_columns <- c("Alley","PoolQC","Fence", "MiscFeature")
training <- training %>% select(-all_of(na_columns))

# Obtaining the remaining number of NA values
ncol(training %>% filter(if_any(everything(), is.na)))
```

There are still

```{r}
# Obtaining the remaining number of NA values
nrow(training %>% filter(if_any(everything(), is.na)))
```

We can see that 877 observations have NA values, therefore we will now concentrate on replacing the missing values, taking into account their importance and if the missing variable is categorical or not.
